---
title: "BIDA302: Lesson 2"
subtitle: "Variable types"
output: learnr::tutorial
runtime: shiny_prerendered
---



<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. -->


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)
library(readxl)

```


## 1. Variable types

In R, data can be stored in a variety of types. For this tutorial, we will focus on numeric and logical variables.


### Variable types in R

| abbreviation | type | example  |
|---|------|---|
| **numeric** |  |  |
|`int` | integers | `31L`  | 
|`dbl`  | doubles, also known as real numbers | `31.5` |
| **data and time** |  |
|`date` | dates | "2020-10-31" |
|`dttm` | date-times (a date + a time) |
| **other** |  |
|`chr`  | character vectors, also known as strings |
|`fctr` | factors, which R uses to represent categorical variables with fixed possible values |
|`lgl` | logical  | `TRUE`, `FALSE`, `NA` |

**Note:** This is an incomplete list! There are other specialized variable types which you may encouter in later, more advanced settings. These are the types we will be working with in this course.


### Creating numeric variables

Numeric variables are at the centre of most data analysis. In R, numeric variables are stored in two ways: as integers (without decimals), or doubles (from the term "double precision", or [double-precision floating-point](https://en.wikipedia.org/wiki/Double-precision_floating-point_format), a computer number format). 

Doubles get used when you need a number with a decimal point or a number bigger than roughly two billion. [^Mersenne_footnote] 


The downside to doubles is that they are imprecise.

```{r numeric_1_example, exercise=TRUE, exercise.eval=FALSE}
# calculate the square root of two, to the 2nd power
x <- sqrt(2) ^ 2
x

# what is x - 2?
x - 2
```

Which isn't precise, but for most application is close enough!

When you are creating a number vector in R, the default state is a double:

```{r numeric_2_example, exercise=TRUE, exercise.eval=FALSE}
d <- 1
typeof(d)

# in a data table we see the variable type shown under the variable name
x <- tribble(~x, 1)
x

```



### Logical variables


`lgl` stands for logical, vectors that contain only TRUE, FALSE, or NA.

Some functions return a logical value; for example:

```{r logical_1_example, exercise=TRUE, exercise.eval=FALSE}
x <- 2
x == 2
```

It's also possible to create a list with logical values:

```{r logical_2_example, exercise=TRUE, exercise.eval=FALSE}
logical_list <- c(TRUE, TRUE, FALSE)
logical_list
```




### Missing values

`NA` are explicitly missing values.

* they can be included as any type: e.g. numeric or character

```{r missing_1_example, exercise=TRUE, exercise.eval=FALSE}

logical_list <- c(TRUE, NA, FALSE)
logical_list

number_list <- c(1, NA, 3)
number_list
```



## 2. Missing Values


### a short example

`is.na` and `anyNA(x)` functions are logical


### Your Turn 1.1: Find the `NA` values

`is.na` and `anyNA(x)` functions are logical

What does `is.na(x)` return? 


```{r missing1, exercise=TRUE, exercise.eval=FALSE}
number_list <- c(1, NA, 3)

# answer
is.na(number_list)

```


There are three values in `number_list`, so `is.na()` runs three tests--only the second one is `NA`.

What about `anyNA(number_list)`?

```{r missing2, exercise=TRUE, exercise.eval=FALSE}
# example
number_list <- c(1, NA, 3)

```



```{r missing2-solution}
number_list <- c(1, NA, 3)

# answer
anyNA(number_list)

# There are three values in `x`, so three tests--only the second one is `NA`

```




What if we put an exclamation mark -- the "not" symbol -- in front of `is.na()`? How does it differ from `is.na()`?

```{r missing3, exercise=TRUE, exercise.eval=FALSE}
# example
number_list <- c(1, NA, 3)

```



```{r missing3-solution}
# example
number_list <- c(1, NA, 3)

!is.na(number_list)
```


## Calculating with `NA`


Run this code to see what happens when there are `NA` values in a variable, and you try to sum them.

Use `na.rm` within the `sum()` function to calculate the sum of `number_list`.


```{r missing4, exercise=TRUE, exercise.eval=FALSE}
# example
number_list <- c(1, NA, 3)

# example
sum(number_list)

```



```{r missing4-solution}

# answer
number_list <- c(1, NA, 3)
sum(number_list, na.rm = TRUE)


```



### Your Turn 1.2: 

First step: review the structure of the `mpg` data set:

```{r missing1_2_1, exercise=TRUE, exercise.eval=FALSE}
mpg
```


Run to create the `displ_class_by_cyl` table:

* group the cars by class and cylinder size, and 

* show the mean displacement (engine size)

```{r missing1_2_2, exercise=TRUE, exercise.eval=FALSE}
# summary table - class by cylinder
displ_class_by_cyl <- mpg %>%
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) %>%
  arrange(desc(displ_mean)) %>%
  pivot_wider(names_from = cyl, values_from = displ_mean) %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

#saveRDS(displ_class_by_cyl, file = "tutorials/BIDA302_solutions/hands-on_2/displ_class_by_cyl.rds")

```

```{r, include=FALSE}
#load displ_class_by_cyl table (since the output of one chunk doesn't carry forward)

displ_class_by_cyl <- readRDS(file = "displ_class_by_cyl.rds")


```


Calculate the mean of `displ_mean` -- note that there are `NA` values.

```{r missing1_2_3, exercise=TRUE, exercise.eval=FALSE}
# example
mean(displ_class_by_cyl$displ_mean)


```



```{r missing1_2_3-solution}


# solution
mean(displ_class_by_cyl$displ_mean, na.rm = TRUE)

```


An alternative solution: use a filter with `!na` to remove the records with `NA` values:

```{r missing1_2_4, exercise=TRUE, exercise.eval=FALSE}
# example
displ_class_by_cyl %>%
  summarise(displ_mean_all = mean(displ_mean))
```



```{r missing1_2_4-solution}


# solution
displ_class_by_cyl %>%
  filter(!is.na(displ_mean)) %>%
  summarise(displ_mean_all = mean(displ_mean))

```



## Summarize with `group()` and `ungroup()`


You'll notice in #1.2 that when we summarize `displ_class_by_cyl` it gives the mean values by class, even though we didn't use any grouping variable.

This is because when we ran the code to create the `displ_class_by_cyl` table, we grouped by `class` and `cyl`. Running the `summarise()` function is applied, it removes one level of the grouping (in this case, `cyl`): 

```{r missing1_2_5, exercise=TRUE, exercise.eval=FALSE}

# example
displ_class_by_cyl

displ_class_by_cyl %>%
  filter(!is.na(displ_mean)) %>%
  summarise(displ_mean_all = mean(displ_mean))


```


If you want the mean of all the values, you have to use `ungroup()` before `summarise()`, to "peel off" `class`


```{r missing1_2_6, exercise=TRUE, exercise.eval=FALSE}

# solution
displ_class_by_cyl %>%
  filter(!is.na(displ_mean)) %>%
  ungroup() %>%
  summarise(displ_mean_all = mean(displ_mean))


```


## Take aways

* using functions to identify unknown values (`NA`) in a variable
* finding `NA` values
* removing or avoiding `NA` values in calculations




***

https://stackoverflow.com/questions/63675677/include-a-chunk-in-an-r-markdown-footnote

[^Mersenne_footnote]:

The maximum positive value for a 32-bit binary integer is 2,147,483,647. This number is tied to the how numbers are stored in 32-bit computers, and is represented as (2^31 - 1). For you number nerds, this happens to be a [double Mersenne prime number](https://en.wikipedia.org/wiki/Double_Mersenne_number).

```{r}
# a Mersenne prime is a prime number that can be represented as "2^n - 1"
2^31-1

# a double Mersenne prime a Mersenne prime is where "n" is also a Mersenne prime
# - in this case, 31 is a Mersenne prime (it's a prime number, and is 2^5 - 1)
2^(2^5-1)-1 

```



