---
title: "ESDL300: Transform Data"
subtitle: "hands-on examples, with answers"
output: learnr::tutorial
runtime: shiny_prerendered
---


<!-- This file by Charlotte Wickham (with modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio and https://github.com/cwickham/data-science-in-tidyverse-solutions. 

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)

-->


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)

```

## Transform Data 1a: manipulate cases

> Please refer to [_R for Data Science_, Data transformation](https://r4ds.had.co.nz/transform.html) for more information about this section.

When working with data, particularly large data sets, you encounter a situation where you need to: 

* subset the data so that it contains only those _records_ (rows) you are interested in (Why deal with a data table that has every province in Canada when you just need the British Columbia data?)

* subset the data so that it contains only those _variables_ (columns) you are interested in (Why keep looking at variables that do not factor into your analysis, or you will not print in your report?)

* create new variables, often through calculations based on variables in your data



To achieve these three goals, we will be using functions from the tidyverse package {dplyr}. The functions are verbs that describe what they do. Here's a list of the {dplyr} functions that you'll be using:


|function |action |
| :--- | :--- |
| `filter()`    | *keep rows*|
| `select()`    | *keep variables (or drop them -var)*|
| `mutate()`    | *create a new variable*|
| `arrange()`    | *sort from smallest to largest*|
| `arrange(desc())`    | *sort from largest to smallest*|



### other operators

In addition to these functions, R has a number of operators that add significantly to your code.

**Arithmetic operators**

| Operator | Description |
| :--- | :--- |
| + | *plus* |
| - | *minus* |
| \* | *multiplication*  |
| / | *division* |
| ^ | *exponential* |

<P>

**Boolean operators** are a special type of operator that return TRUE or FALSE.  

| Operator | Description |
| :--- | :--- |
| ==    | *equal, tests equality* |
| !=    | *not equal, tests inequality* |
| \>    | *greater than, tests greater than* (also >=) |
| <     | *less than, tests less than* (also <=) |
| %in%  | *contains, tests inclusion* |

<P>

Boolean operators can be combined with *and* or *or*.  

| Operator | Description |
| :--- | :--- |
| `&`     | *and, returns true if preceding and following are _both_ true, else false* |
| &#124;  | *or, returns true if _either_ preceding and following are true, else false* |


* see [_R for Data Science_, "Comparisions""](https://r4ds.had.co.nz/transform.html#comparisons) and ["Logical Operators"](https://r4ds.had.co.nz/transform.html#logical-operators).]




### gapminder data

We will be using the data in the {gapminder} package. This data frame has the following variables for each of 142 countries:

* life expectancy, 

* GDP per capita, and 

* population, 

every five years, from 1952 to 2007, as well as the continent for each country.

Take a quick look at the data:


```{r gapminder_data_view, exercise=TRUE}

gapminder

```



## Your Turn 1.1: filter() rows

The `filter()` function allows us to specify the observations (records) that we want to keep in our data analysis.

The general structure of a `filter()` function is:

`filter(.data, variable <test> value)`


See if you can use the logical operators to edit the code below to show:

* the data for Canada

```{r filter1_example, exercise=TRUE, exercise.eval=FALSE}
# example
filter(gapminder, country == "New Zealand")

```

```{r filter1_example-solution}
# answer
filter(gapminder, country == "Canada")

```



* All data for countries in the continent of Oceania

```{r filter2_example, exercise=TRUE, exercise.eval=FALSE}
# example
filter(gapminder, country == "New Zealand")

```


```{r filter2_example-solution}

# answer
filter(gapminder, continent == "Oceania")
```



* Rows where the life expectancy is greater than 82

```{r filter3_example, exercise=TRUE, exercise.eval=FALSE}
# example
filter(gapminder, country == "New Zealand")
```

```{r filter3_example-solution}

# answer
filter(gapminder, lifeExp > 82)

```


## Your Turn 1.2: multiple criteria

### Introducing the pipe operator ` %>% `

(see [_R for Data Science_, Pipes](https://r4ds.had.co.nz/pipes.html) for a deep dive into pipes)

The pipe operator ` %>% ` allows us to put the data first, and then have other functions work with the outcome of the first line.

For example, we could write

`function(data, arg)` or, with a pipe

`data %>% function(arg)` 

If we want to do anything with the first statement, we have to assign it to an intermediate object, and then use that object in the second statement. With the pipe, we can simply add another pipe and the results of the first function are passed to the second statement.

As we get to increasingly complex statements, the power of this will become more evident.

***

Use Boolean operators to alter the code below to return only the rows that contain:

* United States before 1970

```{r filter4_example, exercise=TRUE, exercise.eval=FALSE}
# example 
gapminder %>%                                       # Note use of `%>%` pipe symbol
  filter(country == "New Zealand", year > 2000)

```

```{r filter4_example-solution}
# answer
gapminder %>%
  filter(country == "United States", year < 1970)

# alternative answer
gapminder %>%
  filter(country == "United States" & year < 1970)  # Note use of `&` rather than comma

```

*  Countries where life expectancy in 2007 is below 50 or over 75

```{r filter5_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r filter5_example-solution}
# answer
gapminder %>%
  filter(year == 2007, lifeExp < 50 | lifeExp > 75) # Note that you have to state `lifeExp` twice, as
                                                    # there are two comparisons.
                                                    # What happens if you replace the comma after 2007 with `&`?
                                                    # - Why do you think this might be happening?

# alternate answer
gapminder %>%
  filter(year == 2007) %>% 
  filter(lifeExp < 50 | lifeExp > 75)  # This solution treats each part of the "test" separately


```


### Creating lists

R allows us to create lists using the `c()` function.

A list can be referenced in a `filter()` statement using the `%in%` as the evaluator.

In this example, we filter to get three countries, "New Zealand", "Canada", and "United States".

Note that to get all three, the statement is _OR_, not _AND_! There are no cases where `country == "New Zealand"` and simultaneously `country == "Canada"` ... 

```{r filter6a_example, exercise=TRUE, exercise.eval=FALSE}

gapminder %>% 
  filter(country == "New Zealand" |
           country == "Canada" |
           country == "United States")

```

Instead of this rather verbose syntax, we can achieve the same result by creating a list and using the `%in%` operator to test within the list. 


```{r filter6b_example, exercise=TRUE, exercise.eval=FALSE}

# answer
gapminder %>%
  filter(country %in% c("New Zealand", "Canada", "United States"))

```

Another option would be to create a named object that contains our list, and then use the named object in our `filter()`


```{r filter7a_example, exercise=TRUE, exercise.eval=FALSE}

  
# alternate answer
country_list <- c("New Zealand", "Canada", "United States")   # create object that is list of countries

gapminder %>%
  filter(country %in% country_list)                           # filter uses that object 
  

```



* Create a list object and return records for any of "Egypt", "Spain", "Denmark", "Lebanon" 


```{r filter6_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r filter6_example-solution}
# answer
gapminder %>%
  filter(country %in% c("Egypt", "Spain", "Denmark", "Lebanon"))
  
# alternate answer
country_list <- c("Egypt", "Spain", "Denmark", "Lebanon")   # create object that is list of countries

gapminder %>%
  filter(country %in% country_list)                           # filter uses that object 
  

```

* Return the records for all continents except Europe

```{r filter7_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r filter7_example-solution}
# answer
gapminder %>%
  filter(continent != "Europe")
  
# answer with a double-check
gapminder %>%
  filter(continent != "Europe") %>%
  distinct(continent)                 # the `distinct()` function returns the list of all the 
                                      # continent names after the filter is applied

```

## Your Turn 1.3: sorting with `arrange()`

Sort the records so that the country / year combination with the smallest population are first.

```{r arrange1_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r arrange1_example-solution}

#answer
gapminder %>%
  arrange(pop)

```

Sort the records so that the largest GDP per capita is shown first.



```{r arrange2_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r arrange2_example-solution}

#answer
gapminder %>% arrange(desc(gdpPercap))

#answer
arrange(gapminder, desc(gdpPercap))

#answer
gapminder %>% arrange(-gdpPercap)


```




## Your Turn 1.4: creating new variables

Use the `mutate()` function to calculate a new variable with total GDP.


```{r mutate1_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r mutate1_example-solution}

# answer
gapminder %>%
  mutate(totalGDP = pop * gdpPercap)


```



## Take aways

* Extract cases with `filter()`  
* Filter using a list with `%in%`
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  







## Your Turn 1.5: select() for variables

{gapminder} again


Using `select()` to reduce the number of variables to `country`, `year`, and `lifeExp`:

```{r yourturn4_1_1, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn4_1_1-solution}
# answer
gapminder %>%                                
  select(country, year, lifeExp) 

```



You can use the minus sign to drop variable--that is, to define ones to _not_ select. 

In this example, note we can use `c()` to create a list:


```{r yourturn4_1_2, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn4_1_2-solution}
# answer
gapminder %>%                                
  select(-(c(continent, pop, gdpPercap))) 

# alternate answer
gapminder %>%                                
  select(-continent, -pop, -gdpPercap) 

```



You can also define a range of columns to select by using the `:`

```{r yourturn4_1_3, exercise=TRUE, exercise.eval=FALSE}
# answer
gapminder %>%
  select(country, year:pop)
```


## Take aways

* Restructure data frames with `select()` 
* Different ways to define what variables are selected by `select()`
* Reinforce other data transformation functions





## Complete list of functions


Here's a list of the {dplyr} and {tidyr} functions that are used in this tutorial, and a few more that are handy from time to time:


|function |action |
| :--- | :--- |
| **{dplyr}  | - |
| `filter()`    | *keep rows*|
| `select()`    | *keep variables (or drop them -var)*|
| `mutate()`    | *create a new variable*|
| `distinct()`  | *returns only rows that are unique*| 
| `case_when()` | *is used for "recoding" variable, often used with mutate()*|
| `arrange()`    | *sort from smallest to largest*|
| `arrange(desc())`    | *sort from largest to smallest*|



## Reference

This tutorial is written using the {learnr} package

* github repo: https://github.com/MonkmanMH/tutormanner

* {learnr} reference page: https://rstudio.github.io/learnr/index.html


**License:**

This file by Charlotte Wickham (with modifications by Martin Monkman) is licensed under a [Creative Commons Attribution 4.0 International License.](https://creativecommons.org/licenses/by/4.0/), adapted from the original work at https://github.com/rstudio/master-the-tidyverse by RStudio and https://github.com/cwickham/data-science-in-tidyverse-solutions. 

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)

