---
title: "ESDL300: Introduction to R"
subtitle: "hands-on examples, with answers"
output: learnr::tutorial
runtime: shiny_prerendered
---



<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)
-->


```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.completion = FALSE)

knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)

```

## Objectives

At the end of this tutorial, you will have:

* An understanding of descriptive statistics and exploratory data analysis (EDA)

* The ability to generate descriptive statistics and apply EDA methods using R



## Exploratory Data Analysis (EDA)

So far we've only looked at objects that have one item in them. But most of your data analysis is going to use a data frame (also known as a data table) of some sort. Think of a typical spreadsheet of numbers, with the data in columns and rows.

We will delve into what makes a good data frame structure later in this course, but for now let's practice with some small examples so you can see what's happening. 

In this example, we have a data frame called "anscombe_1" with 2 variables, and each variable has 11 cases in each. Let's take a quick look so we get a sense of the structure:



```{r E1}

anscombe_1 <- anscombe %>% 
  select(x1, y1)

```



```{r E1-1, exercise=TRUE, exercise.setup = "E1"}

anscombe_1

```


We can calculate the mean value of `x1` using the following code. The elements in the code are:

`function(dataframe$variable)`

Note that the dollar sign separates the data frame name from the variable in that data frame. (Later in the course we will see other ways to write the same code.)


```{r E1-2, exercise=TRUE, exercise.setup = "E1"}

mean(anscombe_1$x1)


```


### Your Turn E.1: functions on data frames

Using the `anscombe_1` data frame, write the script to calculate:

* the minimum value of `x1`

* the median value of `y1`


```{r E1-3_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "E1"}


```



```{r E1-3_yt-solution, exercise.setup = "E1"}
# answer

min(anscombe_1$x1)

# answer: 4

median(anscombe_1$y1)

# answer: 7.58


```



### Some two-variable statistical functions

Earlier we saw functions that calculate summary statistics on a single variable, such as `mean()` and `sd()`. R also gives us functions to calculate the statistics on two variables, such as the correlation and the covariance.


**Two-variable statistical functions**

| Function | Description |
| :--- | :--- |
| `cor(x, y)`  | *calculate the correlation between x and y* |
| `cov(x, y)`  | *calculate the covariance between x and y* |

<P>

And again, the statistical/data science foundation of R means that these and many more functions are built in to base R.


### Your Turn E.2: functions on two variables

Using the `anscombe_1` data frame, write the script to calculate:

* the correlation between `x1` and `y1`


```{r E2-1_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "E1"}


```



```{r E2-1_yt-solution, exercise.setup = "E1"}
# answer

cor(anscombe_1$x1, anscombe_1$y1)

# answer: 0.8164205

```


## Anscombe plot



## Summary tables with {dplyr}


## EXERCISE: {dplyr} functions

Use the {dplyr} functions to create a summary table with the population and GDP by continent for the year 1952  


```{r dplyr1_exercise, exercise=TRUE, exercise.eval=FALSE}

```


```{r dplyr1_exercise-solution}

# answer
gapminder %>%
  filter(year == 1952) %>%               # filter to get the records we need
  mutate(totalGDP = pop * gdpPercap) %>% # note that we need to add the new variable calculated above!
  group_by(continent) %>%                # this defines the grouping category
  summarize(pop = sum(pop),              # 
            totalGDP = sum(totalGDP))    #
```




## Your Turn 1.5

Alter the code to add a `prev_lifeExp` column that contains the life expectancy from the previous record.

(Hint: refer to the ["Data Transformation"](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) cheatsheet; you want to offset elements by one)


```{r dplyr2_exercise, exercise=TRUE, exercise.eval=FALSE}

```


```{r dplyr2_exercise-solution}

# answer
gapminder %>%
  mutate(prev_lifeExp = lag(lifeExp))

# challenge hint: 
# - note that there is an `NA` in the first row, since there's no prior row to pull from
```



```{r quiz1_5}
quiz(
  question("Why isn't this quite the 'life expectency five years ago'?",
    answer("The `lag()` function doesn't know when the country changes so is pulling the last record for the previous country", correct = TRUE),
    answer("The `lag() function is only 1 step, not 5, so it's only one year apart"),
    answer("The data are every 4 years apart"),
    answer("The question is a trick--the value _is_ 'life expectency five years ago'")
  )
)
```



## Transform Data 2: summarize

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

When working with data, one of the first things in our analysis is often the calculation of summary statistics. We may want the number of cases, or the maximum value, or the mean of all the records. 

Here's a list of the {dplyr} functions that you'll be using:



|function |action |
| :--- | :--- |
| `group_by()`  | *declare subsets in data*|
| `summarize()` | *summarize the data, by groups if they have been declared*|
| `tally()`     | *counting (by groups if group_by() applied)*|


**gapminder**

Take a quick look at the data in the {gapminder} package.

```{r yourturn2_0, exercise=TRUE, exercise.eval=FALSE}
gapminder
```



## Your Turn 2.1

Use `summarise()` to compute three statistics about the data:

* The first (minimum) year in the dataset
* The last (maximum) year in the dataset
* The number of unique countries

```{r yourturn2_1, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_1-solution}
# answer
gapminder %>%
  summarize(year_min = min(year),
            year_max = max(year),
            n_countries = n_distinct(country))

```

## Your Turn 2.2

Extract the rows for African countries in 2007.

Then find:

1. The number of unique countries
2. The median life expectancy of African countries as a group

```{r yourturn2_2, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_2-solution}
# answer
gapminder %>% 
  filter(continent == "Africa", year == 2007) %>% 
  summarise(n_countries = n_distinct(country),
    med_life_exp = median(lifeExp))


```


## Your Turn 2.3

Find the median life expectancy by continent in 2007.

```{r yourturn2_3, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_3-solution}
gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>% 
  summarise(med_life_exp = median(lifeExp)) 

```


## Your Turn 2.4

Find the median life expectancy by continent for each year in the dataframe.

```{r yourturn2_4, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_4-solution}

gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp))

```


## Your Turn 2.5


Find the country with biggest jump in life expectancy (between any two consecutive records) for each continent.

```{r yourturn2_5, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_5-solution}
# One of many solutions
gapminder %>%
  group_by(country) %>%
  mutate(prev_lifeExp = lag(lifeExp),
         jump = lifeExp - prev_lifeExp) %>%
  arrange(desc(jump))


```


## Take aways

* Make tables of summaries with `summarise()`  
* Do groupwise operations with `group_by()`

***



## Transform Data 3: pivot

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License -->

Pivoting the data in your file can be very helpful in your analysis. It's particularly helpful when your data is not "tidy". For example, there might be 

* One variable spread across multiple columns

* One observation scattered across multiple rows.

You might also want to pivot the data to make it more human readable (a more compact layout that's easily scanned) or machine readable.

See [_R for Data Science_, Tidy data: Pivoting](https://r4ds.had.co.nz/tidy-data.html#pivoting)


The functions to pivot your data table are in the package {tidyr}.


|function |action |
| :--- | :--- |
| **{tidyr}  | - |
| `pivot_wider()` | Pivot data from long to wide |
| `pivot_longer()` | Pivot data from wide to long |



## Your Turn 3.1

A pivot table crosstab using the {mpg} data package of automobile fuel economy.

First, a look at the source table.

```{r yourturn3_0, exercise=TRUE, exercise.eval=FALSE}

mpg
```

Using `group_by` and `summarise` to create a summary table of the average engine displacement by vehicle class:

```{r yourturn3_1_1, exercise=TRUE, exercise.eval=FALSE}

mpg %>%                                
  group_by(class) %>%                   
  summarise(displ_mean = mean(displ)) 

```

Now do the same with vehicle class and number of cylinders:

```{r yourturn3_1_2, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_2-solution}
# answer
class_by_cyl <- mpg %>%                 # note that an object `class_by_cyl` is created
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) 

class_by_cyl                            # in order to print the object, just name it


```


What happens if you reverse the order of the grouping variables?

```{r yourturn3_1_3, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_3-solution}
# answer
class_by_cyl <- mpg %>%                 # note that an object `class_by_cyl` is created
  group_by(cyl, class) %>%
  summarise(displ_mean = mean(displ)) 

class_by_cyl                            # in order to print the object, just name it

```


Create a pivot (crosstab) table with class by number of cylinders:

```{r, include=FALSE}
#load class_by_cyl table (since the output of one chunk doesn't carry forward)

#saveRDS(class_by_cyl, file = "tutorials/BIDA302_solutions/data/class_by_cyl.rds")
class_by_cyl <- readRDS(file = "data/class_by_cyl.rds")


```


```{r yourturn3_1_4, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_4-solution}
# answer
# pivot table (wide)

class_by_cyl_pivot <- class_by_cyl %>%
  pivot_wider(names_from = cyl, values_from = displ_mean)

class_by_cyl_pivot


```

Now, unpivot it back to the original structure:

```{r, include=FALSE}
#saveRDS(class_by_cyl_pivot, file = "tutorials/BIDA302_solutions/data/class_by_cyl_pivot.rds")


#load class_by_cyl_pivot table (since the output of one chunk doesn't carry forward)

class_by_cyl_pivot <- readRDS(file = "data/class_by_cyl_pivot.rds")


```


```{r yourturn3_1_5, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_5-solution}
# and back to longer...

displ_class_by_cyl <- class_by_cyl_pivot %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```




```{r quiz3_1}
quiz(
  question("What do you notice about the structure of the unpivoted table?",
    answer("It's exactly the same as it was before"),
    answer("The variables (columns) are in different order"),
    answer("There are now `NA` values", correct = TRUE),
    answer("The values in the cells have changed")
  )
)
```





## Your Turn 3.2

Making some summary tables with the {gapminder} data.

Create a table where:

* object name `lifeExp_by_continent` that has 
* the median life expectency for each continent for every year in the dataframe, 
* where the continent is the column and the year is the row.

(This is a restructuring of the table from 2.4)

```{r yourturn3_2_1, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_2_1-solution}
# answer

lifeExp_by_continent <- gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp)) %>%
  pivot_wider(names_from = continent, values_from = median_lifeExp)

lifeExp_by_continent

```

## Your Turn 3.3

Create a table that shows:

* GDP per capita for
* the countries Canada, United States, and Mexico are the columns, and 
* the years after 1980 are the rows

```{r yourturn3_3_1, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_3_1-solution}

gapminder %>%
  filter(year > 1980,
         country %in% c("Canada", "United States", "Mexico")) %>%
  select(country, year, gdpPercap) %>%        # select() is covered in section 4
  pivot_wider(names_from = country, values_from = gdpPercap)

```



## Take aways

* Restructure data tabulations with `pivot_wider()` and `pivot_longer`  



## Complete list of functions


Here's a list of the {dplyr} and {tidyr} functions that are used in this tutorial, and a few more that are handy from time to time:


|function |action |
| :--- | :--- |
| **{dplyr}  | - |
| `filter()`    | *keep rows*|
| `select()`    | *keep variables (or drop them -var)*|
| `mutate()`    | *create a new variable*|
| `group_by()`  | *declare subsets in data*|
| `summarize()` | *summarize the data, by groups if they have been declared*|
| `rename()`    | *renaming variables*|
| `tally()`     | *counting (by groups if group_by() applied)*|
| `distinct()`  | *returns only rows that are unique*| 
| `case_when()` | *is used for "recoding" variable, often used with mutate()*|
| `arrange()`    | *sort from smallest to largest*|
| `arrange(desc())`    | *sort from largest to smallest*|
| **{tidyr}  | - |
| `pivot_wider()` | |
| `pivot_longer()` | |



## Reference

This tutorial is written using the {learnr} package

* github repo: https://github.com/MonkmanMH/tutormanner

* {learnr} reference page: https://rstudio.github.io/learnr/index.html


**License:**

This file by Martin Monkman is licensed under a [Creative Commons Attribution 4.0 International License.](https://creativecommons.org/licenses/by/4.0/)

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)




## E. Data frames

So far we've only looked at objects that have one item in them. But most of your data analysis is going to use a data frame (also known as a data table) of some sort think of a typical spreadsheet of numbers, with the data in columns and rows.

We will delve into what makes a good data frame structure later in this course, but for now let's practice with some small examples so you can see what's happening. 

In this example, we have a data frame called "anscombe_1" with 2 variables, and each variable has 11 cases in each. Let's take a quick look so we get a sense of the structure:



```{r E1}

anscombe_1 <- anscombe %>% 
  select(x1, y1)

```



```{r E1-1, exercise=TRUE, exercise.setup = "E1"}

anscombe_1

```


We can calculate the mean value of `x1` using the following code. The elements in the code are:

`function(dataframe$variable)`

Note that the dollar sign separates the data frame name from the variable in that data frame. (Later in the course we will see other ways to write the same code.)


```{r E1-2, exercise=TRUE, exercise.setup = "E1"}

mean(anscombe_1$x1)


```


### Your Turn E.1: functions on data frames

Using the `anscombe_1` data frame, write the script to calculate:

* the minimum value of `x1`

* the median value of `y1`


```{r E1-3_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "E1"}


```



```{r E1-3_yt-solution, exercise.setup = "E1"}
# answer

min(anscombe_1$x1)

# answer: 4

median(anscombe_1$y1)

# answer: 7.58


```



### Some two-variable statistical functions

Earlier we saw functions that calculate summary statistics on a single variable, such as `mean()` and `sd()`. R also gives us functions to calculate the statistics on two variables, such as the correlation and the covariance.


**Two-variable statistical functions**

| Function | Description |
| :--- | :--- |
| `cor(x, y)`  | *calculate the correlation between x and y* |
| `cov(x, y)`  | *calculate the covariance between x and y* |

<P>

And again, the statistical/data science foundation of R means that these and many more functions are built in to base R.


### Your Turn E.2: functions on two variables

Using the `anscombe_1` data frame, write the script to calculate:

* the correlation between `x1` and `y1`


```{r E2-1_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "E1"}


```



```{r E2-1_yt-solution, exercise.setup = "E1"}
# answer

cor(anscombe_1$x1, anscombe_1$y1)

# answer: 0.8164205

```


