---
title: "python in learnr: readme"
output: html_notebook
---

Some notes to keep track of what I have done.

## setting up the rstudio.cloud environment

This builds on the base project for BIDA302.

The installed R packages are:

```{r tidyverse}
# tidyverse
install.packages("tidyverse")
install.packages("readxl")

# data
install.packages("gapminder")
install.packages("nycflights13")
install.packages("palmerpenguins")

```

The project also contains the necessary elements to run Python and Pandas:

```{r}
install.packages("reticulate")

reticulate::py_config()
reticulate::py_install("pandas")

```

Similar to an R `library()` function call, you also need to "import pandas" in a python chunk

```{python}
# import the `pandas` module so that it's available for use
import pandas as pd

```

That's it!


## Reading a CSV into the {learnr} environment

(notes in DM to @CMastication)

>It seems that the bug I am running into is directly related to the reading of a csv file into the {learnr} environment ... it's happening in the R version of the same code. I won't have time to tackle any further changes until the evening (I have a full afternoon schedule with the day job!)
Thanks again for asking the question and giving me motivation to get on with this!
Update: I got it running locally (see updates on the github page), but it bonks out in http://rstudio.cloud (too much behind the scenes processing in running the R markdown to {learnr}?) and fails completely when I try to push it to http://shinyapps.io (my hypothesis is that publishing there doesn't bundle a python version, so fails)
Turns out the big bug also applies to R {learnr} packages--any data import can't be from the same directory as the Rmd. That works great for a straight-up Rmd, but the html rendering seems to be looking in the wrong place. I solved this by parking the csv in the github repo, and then pointing there in the "setup" chunk
Just realized I should copy this and write a blog post and/or RStudio Community entry!

Next hypothesis: the source CSV is too big for rstudio.cloud. Let's try a sample!

```{r}
library(dplyr)

set.seed(42)

flights_10pct <- nycflights13::flights %>% 
  sample_frac(0.1)

readr::write_csv(flights_10pct, "flights_10pct.csv")

```


