---
title: "BIDA302: Lesson 1"
subtitle: "hands-on examples, with answers"
output: learnr::tutorial
runtime: shiny_prerendered
---



<!-- This file by Charlotte Wickham (with modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio and https://github.com/cwickham/data-science-in-tidyverse-solutions. 

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)

-->


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)

```

## Transform Data 1: manipulate cases

For this section, we will be using some of the functions from the tidyverse package {dplyr}. The functions are verbs that describe what they do. Here's a list of the {dplyr} functions that you'll be using:


|function |action |
| :--- | :--- |
| `filter()`    | *keep rows*|
| `mutate()`    | *create a new variable*|
| `group_by()`  | *declare subsets in data*|
| `select()`    | *keep variables (or drop them -var)*|
| `summarize()` | *summarize the data, by groups if they have been declared*|
| `rename()`    | *renaming variables*|
| `tally()`     | *counting (by groups if group_by() applied)*|
| `distinct()`  | *returns only rows that are unique*| 
| `case_when()` | *is used for "recoding" variable, often used with mutate()*|


### other operators

**Arithmetic operators**

| Operator | Description |
| :--- | :--- |
| + | *plus* |
| - | *minus* |
| \* | *multiplication*  |
| / | *division* |
| ^ | *exponential* |

<P>

**Boolean operators** are a special type of operator that return TRUE or FALSE.  

| Operator | Description |
| :--- | :--- |
| ==    | *equal, tests equality* |
| !=    | *not equal, tests inequality* |
| \>    | *greater than, tests greater than* (also >=) |
| <     | *less than, tests less than* (also <=) |
| %in%  | *contains, tests inclusion* |

<P>

Boolean operators can be combined with *and* or *or*.  

| Operator | Description |
| :--- | :--- |
| `&`     | *and, returns true if preceeding and following are both true, else false* |
| &#124;  | *or, returns true if either preceeding and following are true, else false* |


* see [_R for Data Science_, "Comparisions""](https://r4ds.had.co.nz/transform.html#comparisons) and ["Logical Operators"](https://r4ds.had.co.nz/transform.html#logical-operators).]




## gapminder data

We will be using the data in the {gapminder} package. Take a quick look at the data:


```{r gapminder_data_view, exercise=TRUE}

gapminder

```



## Your Turn 1.1: filter() rows

See if you can use the logical operators to manipulate our code below to show:

The data for Canada

```{r filter1_example, exercise=TRUE, exercise.eval=FALSE}
# example
filter(gapminder, country == "New Zealand")

```

```{r filter1_example-solution}
# answer
filter(gapminder, country == "Canada")

```



All data for countries in Oceania

```{r filter2_example, exercise=TRUE, exercise.eval=FALSE}
# example
filter(gapminder, country == "New Zealand")

```


```{r filter2_example-solution}

# answer
filter(gapminder, continent == "Oceania")
```



Rows where the life expectancy is greater than 82
```{r filter3_example, exercise=TRUE, exercise.eval=FALSE}
# example
filter(gapminder, country == "New Zealand")
```

```{r filter3_example-solution}

# answer
filter(gapminder, lifeExp > 82)

```


## Your Turn 1.2: multiple criteria

Use Boolean operators to alter the code below to return only the rows that contain:

* United States before 1970

```{r filter4_example, exercise=TRUE, exercise.eval=FALSE}
# example 
gapminder %>%                                       # Note use of `%>%` pipe symbol
  filter(country == "New Zealand", year > 2000)

```

```{r filter4_example-solution}
# answer
gapminder %>%
  filter(country == "United States", year < 1970)

# alternative answer
gapminder %>%
  filter(country == "United States" & year < 1970)  # Note use of `&` rather than comma

```

*  Countries where life expectancy in 2007 is below 50 or over 75

```{r filter5_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r filter5_example-solution}
# answer
gapminder %>%
  filter(year == 2007, lifeExp < 50 | lifeExp > 75) # Note that you have to state `lifeExp` twice, as
                                                    # there are two comparisons.
                                                    # What happens if you replace the comma after 2007 with `&`?
                                                    # - Why do you think this might be happening?



```



* Create a list object and return records for any of "New Zealand", "Canada" or "United States"


```{r filter6_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r filter6_example-solution}
# answer
gapminder %>%
  filter(country %in% c("New Zealand", "Canada", "United States"))
  
# alternate answer
country_list <- c("New Zealand", "Canada", "United States")   # create object that is list of countries

gapminder %>%
  filter(country %in% country_list)                           # filter uses that object 
  

```

* Return records for all continents except Europe

```{r filter7_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r filter7_example-solution}
# answer
gapminder %>%
  filter(continent != "Europe")
  
# answer with a double-check
gapminder %>%
  filter(continent != "Europe") %>%
  distinct(continent)                 # the `distinct()` function returns the list of all the 
                                      # continent names after the filter is applied

```

## Your Turn 1.3: sorting with `arrange()`

Sort the records so that the country / year combinatino with the smallest population are first.

```{r arrange1_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r arrange1_example-solution}

#answer
gapminder %>%
  arrange(pop)

```

Sort the records so that the largest GDP per capita is shown first.



```{r arrange2_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r arrange2_example-solution}

#answer
gapminder %>% arrange(desc(gdpPercap))

#answer
arrange(gapminder, desc(gdpPercap))

#answer
gapminder %>% arrange(-gdpPercap)


```


## Your Turn 1.4: creating new variables

Use the `mutate()` function to calculate a new variable with total GDP.


```{r mutate1_example, exercise=TRUE, exercise.eval=FALSE}

```


```{r mutate1_example-solution}

# answer
gapminder %>%
  mutate(totalGDP = pop * gdpPercap)


```


## EXERCISE: {dplyr} functions

Use the {dplyr} functions to create a summary table with the population and GDP by continent for the year 1952  


```{r dplyr1_exercise, exercise=TRUE, exercise.eval=FALSE}

```


```{r dplyr1_exercise-solution}

# answer
gapminder %>%
  filter(year == 1952) %>%               # filter to get the records we need
  mutate(totalGDP = pop * gdpPercap) %>% # note that we need to add the new variable calculated above!
  group_by(continent) %>%                # this defines the grouping category
  summarize(pop = sum(pop),              # 
            totalGDP = sum(totalGDP))    #
```




## Your Turn 1.5

Alter the code to add a `prev_lifeExp` column that contains the life expectancy from the previous record.

(Hint: refer to the ["Data Transformation"](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) cheatsheet; you want to offset elements by one)

Extra challenge: Why isn't this quite the 'life expectency five years ago'?


```{r dplyr2_exercise, exercise=TRUE, exercise.eval=FALSE}

```


```{r dplyr2_exercise-solution}

# answer
gapminder %>%
  mutate(prev_lifeExp = lag(lifeExp))

# challenge hint: 
# - note that there is an `NA` in the first row, since there's no prior row to pull from
# - look at the `prev_lifeExp` when the country changes from Afghanistan to Albania 
```


## Take aways

* Extract cases with `filter()`  
* Filter using a list with `%in%`
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  

***

## Transform Data 2: summarize

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->


**gapminder**

Take a quick look at the data in the {gapminder} package.

```{r yourturn2_0, exercise=TRUE, exercise.eval=FALSE}
gapminder
```



## Your Turn 2.1

Use summarise() to compute three statistics about the data:

* The first (minimum) year in the dataset
* The last (maximum) year in the dataset
* The number of unique countries

```{r yourturn2_1, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_1-solution}
# answer
gapminder %>%
  summarize(year_min = min(year),
            year_max = max(year),
            n_countries = n_distinct(country))

```

## Your Turn 2.2

Extract the rows for African countries in 2007.

Then find:

1. The number of unique countries
2. The median life expectancy of African countries as a group

```{r yourturn2_2, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_2-solution}
# answer
gapminder %>% 
  filter(continent == "Africa", year == 2007) %>% 
  summarise(n_countries = n_distinct(country),
    med_life_exp = median(lifeExp))


```


## Your Turn 2.3

Find the median life expectancy by continent in 2007.

```{r yourturn2_3, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_3-solution}
gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>% 
  summarise(med_life_exp = median(lifeExp)) 

```


## Your Turn 2.4

Find the median life expectancy by continent for each year in the dataframe.

```{r yourturn2_4, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_4-solution}

gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp))

```


## Your Turn 2.5


Find the country with biggest jump in life expectancy (between any two consecutive records) for each continent.

```{r yourturn2_5, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn2_5-solution}
# One of many solutions
gapminder %>%
  group_by(country) %>%
  mutate(prev_lifeExp = lag(lifeExp),
         jump = lifeExp - prev_lifeExp) %>%
  arrange(desc(jump))


```


## Take aways

* Make tables of summaries with `summarise()`  
* Do groupwise operations with `group_by()`

***



## Transform Data 3: pivot

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License -->


## Your Turn 3.1

A pivot table crosstab using the {mpg} data package of automobile fuel economy.

First, a look at the source table.

```{r yourturn3_0, exercise=TRUE, exercise.eval=FALSE}

mpg
```

Using `group_by` and `summarise` to create a summary table of the average engine displacement by vehicle class:

```{r yourturn3_1_1, exercise=TRUE, exercise.eval=FALSE}

mpg %>%                                
  group_by(class) %>%                   
  summarise(displ_mean = mean(displ)) 

```

Now do the same with vehicle class and number of cylinders:
```{r yourturn3_1_2, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_2-solution}
# answer
class_by_cyl <- mpg %>%                 # note that an object `class_by_cyl` is created
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) 

class_by_cyl                            # in order to print the object, just name it


```


What happens if you reverse the order of the grouping variables?

```{r yourturn3_1_3, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_3-solution}
# answer
class_by_cyl <- mpg %>%                 # note that an object `class_by_cyl` is created
  group_by(cyl, class) %>%
  summarise(displ_mean = mean(displ)) 

class_by_cyl                            # in order to print the object, just name it

```


Create a pivot (crosstab) table with class by number of cylinders:

```{r, include=FALSE}
#load class_by_cyl table (since the output of one chunk doesn't carry forward)

#saveRDS(class_by_cyl, file = "tutorials/BIDA302_solutions/data/class_by_cyl.rds")
class_by_cyl <- readRDS(file = "data/class_by_cyl.rds")


```


```{r yourturn3_1_4, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_4-solution}
# answer
# pivot table (wide)

class_by_cyl_pivot <- class_by_cyl %>%
  pivot_wider(names_from = cyl, values_from = displ_mean)

class_by_cyl_pivot


```

Now, unpivot it back to the original structure:

```{r, include=FALSE}
#saveRDS(class_by_cyl_pivot, file = "tutorials/BIDA302_solutions/data/class_by_cyl_pivot.rds")


#load class_by_cyl_pivot table (since the output of one chunk doesn't carry forward)

class_by_cyl_pivot <- readRDS(file = "data/class_by_cyl_pivot.rds")


```


```{r yourturn3_1_5, exercise=TRUE, exercise.eval=FALSE}

```


```{r yourturn3_1_5-solution}
# and back to longer...

displ_class_by_cyl <- class_by_cyl_pivot %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```




```{r quiz}
quiz(
  question("What do you notice about the structure of the unpivoted table?",
    answer("It's exactly the same as it was before"),
    answer("The variables (columns) are in different order"),
    answer("There are now `NA` values", correct = TRUE),
    answer("The values in the cells have changed")
  )
)
```





## Your Turn 3.2

Making some summary tables with the {gapminder} data.

Create a table where:

* object name `lifeExp_by_continent` that has 
* the median life expectency for each continent for every year in the dataframe, 
* where the continent is the column and the year is the row.

(This is a restructuring of the table from 2.4)

```{r}
# answer

lifeExp_by_continent <- gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp)) %>%
  pivot_wider(names_from = continent, values_from = median_lifeExp)

lifeExp_by_continent

```

## Your Turn 3.3

Create a table that shows:

* GDP per capita for
* the countries Canada, United States, and Mexico are the columns, and 
* the years after 1980 are the rows

```{r}

gapminder %>%
  filter(year > 1980,
         country %in% c("Canada", "United States", "Mexico")) %>%
  select(country, year, gdpPercap) %>%        # select() is covered in section 4
  pivot_wider(names_from = country, values_from = gdpPercap)

```



## Take aways

* Restructure data tabulations with `pivot_wider()` and `pivot_longer`  


***

## Transform data 4: select()

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License -->


```{r}
library(tidyverse)
# data packages
library(gapminder)

```


## Your Turn 4.1

{gapminder} again

```{r}
gapminder
```

Using `select()` to reduce the number of variables:

```{r}
# answer
gapminder %>%                                
  select(country, year, lifeExp) 

```



You can use the minus sign to drop variable--that is, to define ones to _not_ select. 

In this example, note we can use `c()` to create a list:


```{r}
# answer
gapminder %>%                                
  select(-(c(continent, pop, gdpPercap))) 

# alternate answer
gapminder %>%                                
  select(-continent, -pop, -gdpPercap) 

```



You can define a range of columns to select by using the `:`

```{r}
# answer
gapminder %>%
  select(country, year:pop)
```


## Take aways

* Restructure data frames with `select()` and `pull()`  
* Reinforce other data transformation functions



