---
title: "R multiple choice questions"
author: "Martin Monkman"
date: "20/03/2020"
output: learnr::tutorial
runtime: shiny_prerendered
---



<!-- 

This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. 

Original sources are noted.

-->

This file contains a variety of multiple choice questions, formatted for {learnr} tutorials.

The original sources are noted at the top of each section.


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)

```

An assortment of R-oriented multiple choice questions, suitable for {learnr} tutorials

## Base R


### data-flair

Modified from https://data-flair.training/blogs/r-programming-quiz-part-1/

```{r base_01}
quiz(
  question("Which function is used to create a vector with more than one element?",
    answer("`library()`"),
    answer("`plot()`"),
    answer("`c()`", correct = TRUE),
    answer("`seq()`")
  )
)
```



```{r base_02}
quiz(
  question("Which of the following statements returns the maximum value of `x`, excluding missing values?",
    answer("`rm(x)`"),
    answer("`max(x, na.rm=TRUE)`", correct = TRUE),
    answer("`all(x)`"),
    answer("`x %in% y`")
  )
)
```

```{r base_03}
quiz(
  question("R version 1.0.0 was released on which date?",
    answer("1999-12-31"),
    answer("2000-02-29", correct = TRUE),
    answer("2001-01-01"),
    answer("2010-10-10")
  )
)
```


```{r base_04}
quiz(
  question("Attributes of an object (if any) can be accessed using the ______ function.",
    answer("`objects()`"), 
    answer("`attributes()`", correct = TRUE),
    answer("`att()`"),
    answer("`fun`()")
  )
)
```


```{r base_05}
quiz(
  question("Which function would you use to return the names of the objects in the specified enviroment?",
    answer("`list()`"), 
    answer("`ls()`", correct = TRUE),
    answer("`names()`"),
    answer("`fun`()")
  )
)
```





```{r base_06}

quiz(
  question("What would be the output of the following code?  <br>
           `x <- 1:4` <br>
           `y <- 6:9` <br>
           `z <- x + y` <br>
           `z`",
    answer("7 9 11 13", correct = TRUE), 
    answer("7 9 11 13 14"),
    answer("9 11 13"),
    answer("NA")
  ),
  question("What would be the output of the following code?  <br>
           `x <- 1:4` <br>
           `y <- c(6, 7, 8, 9)` <br>
           `z <- x + y` <br>
           `z`",
    answer("7 9 11 13", correct = TRUE), 
    answer("7 9 11 13 14"),
    answer("9 11 13"),
    answer("NA")
  ),
  question("What would be the output of the following code?  <br>
           `x <- 1:4` <br>
           `y <- c(6, 7, NA, 9)` <br>
           `z <- x + y` <br>
           `z`",
    answer("7 9 11 13"), 
    answer("7 9 11 13 14"),
    answer("7 9 NA 13", correct = TRUE),
    answer("NA")
  )
  
)
```



Dealing with NAs

```{r base_07}

quiz(
  question("What would be the output of the following code?  <br>
           `x <- c(1, 2, 3)` <br>
           `sum(x)`",
    answer("1 2 3"), 
    answer("123"),
    answer("6", correct = TRUE),
    answer("NA")
  ),
  question("What would be the output of the following code?  <br>
           `x <- c(1, 2, 3, NA` <br>
           `sum(x)`",
    answer("1 2 3 NA"), 
    answer("123NA"),
    answer("6"),
    answer("NA", correct = TRUE)
  ),
  question("What would be the output of the following code?  <br>
           `x <- c(1, 2, 3, NA` <br>
           `sum(x, na.rm = TRUE)`",
    answer("1 2 NA 4"), 
    answer("12NA4"),
    answer("6", correct = TRUE),
    answer("NA")
  )
)
```



Boolean operators

```{r base_07}

quiz(
  question("What would be the output of the following code?  <br>
           `x <- 1:4` <br>
           `x > 2`",
    answer("FALSE FALSE TRUE TRUE", correct = TRUE), 
    answer("FALSE TRUE TRUE TRUE"), 
    answer("FALSE"), 
    answer("TRUE"), 
    answer("NA")
  ),
  question("What would be the output of the following code?  <br>
           `x <- 1:4` <br>
           `x < 2`",
    answer("TRUE TRUE FALSE FALSE"), 
    answer("TRUE FALSE FALSE FALSE", correct = TRUE), 
    answer("FALSE"), 
    answer("TRUE"), 
    answer("NA")
  )
  
)
```


Q.26 What would be the output of the following code?

> x <- 1:4

> x > 2
FALSE FALSE TRUE TRUE
1 2 3 4
1 2 3 4 5




Next:

https://gist.github.com/peterhurford/010899c596862805dece



## tidyverse

### dplyr

Q.23 Which of the following sort dataframe by the order of the elements in B
a.x[rev(order(x$B)),]
b.x[ordersort(x$B),]
c.x[order(x$B),]


Modified from: https://study.com/academy/practice/quiz-worksheet-tidyverse-in-r-programming.html


Modified from: https://www.r-bloggers.com/test-your-tidyness-a-short-quiz-to-check-your-tidyverse-capabilities/



