---
title: "Plotting in R"
subtitle: "BIDA302: Lesson 3.2"
output: learnr::tutorial
runtime: shiny_prerendered
---


<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. 

Some of these examples are modified from http://euclid.psych.yorku.ca/www/psy6135/tutorials/gapminder.html. Thanks Michael!

-->


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)

library(gapminder)

# load data
gapminder_2007 <- readRDS(file = "gapminder_2007.rds")

```

## 1. Plotting

### Objectives

* Produce scatter plots, histograms, and time series plots using {ggplot2}.

* Set universal plot settings.
 
* Describe what faceting is and apply faceting in a `ggplot()` call.
 
* Modify the aesthetics of an existing plot (including color).
 

### REFERENCES

These two chapters in _R for Data Science_ provide additional background:

* ["Data visualisation"](https://r4ds.had.co.nz/data-visualisation.html)

* ["Exploratory Data Analysis"](https://r4ds.had.co.nz/exploratory-data-analysis.html)




## 1. Scatter plot: GDP by life expectancy

Scatter plots are sometimes known as "X-Y plots" because they compare two values for the same observation. 


For this and some of the other examples, we will be using just the 2007 data from the {gapminder} package.


### Your Turn 1.1

Create a dataframe object with the name "gapminder_2007" by filtering the gapminder data.

```{r scatterplot1, exercise=TRUE, exercise.eval=FALSE}

```

```{r scatterplot1-solution}
# filter the data
gapminder_2007 <- filter(gapminder, year == 2007)
# show the new data object
gapminder_2007

```
<!--
#saveRDS(gapminder_2007, file = "tutorials/BIDA302_solutions/hands-on_3/gapminder_2007.rds")
#gapminder_2007 <- readRDS(file = "gapminder_2007.rds")
-->

### {ggplot} function

Let's take a look at the structure of a basic `ggplot()` function:

`ggplot(data = <DATA>) + `

`  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`


As you can see, the first thing in the function is the data frame you are using. 

Next comes the `geom_` which is the geometric structure you want to use--some examples are a line, or a histogram, or a point. 

We also need to specify which variables to plot; these go into `geom` as the `mapping = aes()`. Note that often, you will see a coding shortcut of just `aes()` (that is, no `mapping =`). 



This code creates a simple scatterplot using the {gapminder} data, showing all the countries in the world for 2007, the per capita GDP on the _x_ axis, and life expectancy on the _y_. 



This uses the `geom_point()` function.

 
```{r scatterplot1_2_1, exercise=TRUE, exercise.eval=FALSE}
# example

ggplot(gapminder_2007, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
  

```


### Your Turn 1.2


Now change the colour of the points by adding the `colour = ` argument to `geom_point()`.

Note that the name of the colour has to be in quotation marks.

```{r scatterplot1_2_2, exercise=TRUE, exercise.eval=FALSE}

```

```{r scatterplot1_2_2-solution}


# change the colour of the points
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(colour = "blue")           


```

Create a variant of the above plot, with each continent coloured differently.

* Note that in this example, we want to add another variable (`continent`) so it gets added to the `aes()` list. 
 
```{r scatterplot1_2_3, exercise=TRUE, exercise.eval=FALSE}

```

```{r scatterplot1_2_3-solution}
# solution
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point()

```

Try it again with shape instead of colour.


```{r scatterplot1_2_4, exercise=TRUE, exercise.eval=FALSE}

```


```{r scatterplot1_2_4-solution}
# solution

ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, shape = continent)) +
  geom_point()

```


## 2. Histogram: life expectancy

Histograms are ways to "bin" continuous variables into categories, and then count the number of observations in each bin.

The geom is `geom_histogram()`.

Note that you don't need a `y =` variable; the Y axis is the count of the number of observations.

### Your turn 2.1

Make a histogram using `gapminder_2007`, showing the count of countries by life expectancy.

```{r histogram2_1, exercise=TRUE, exercise.eval=FALSE}

```

```{r histogram2_1-solution}
# solution
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram()


```


### Your turn 2.2

An important option in the `geom_histogram()` function is `binwidth = `. This sets the range that each bin covers. Experiment with adding bin widths of 3, 5, and 10 to the histogram

```{r histogram2_2_1, exercise=TRUE, exercise.eval=FALSE}

```

```{r histogram2_2_1-solution}
# solution
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 3)


```

Some more formatting...like you did earlier with the points, change the colour of the bars to red.

```{r histogram2_2_2, exercise=TRUE, exercise.eval=FALSE}

```

```{r histogram2_2_2-solution}
# solution
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 3, colour = "red")


```

"colour" is the border of the bar! Use "fill" for the inside -- make a white bar with a black border

```{r histogram2_2_3, exercise=TRUE, exercise.eval=FALSE}

```

```{r histogram2_2_3-solution}
# solution
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 3, fill = "white", colour = "black")


```


### Your turn 2.3

What do you think happens when you put a "fill" option in the `aes()` function?

```{r histogram2_3, exercise=TRUE, exercise.eval=FALSE}

```

```{r histogram2_3-solution}
# solution

# solution
ggplot(gapminder_2007, aes(x = lifeExp, fill = continent)) +
  geom_histogram(binwidth = 3, colour = "black")

```



***

## 3. Line graph: change in life expectancy over time

Line graphs join the dots and are useful for showing change over time.

The geom is `geom_line()`.


### Your turn 3.1

Make a line graph using `gapminder`, showing the change of life expectancy for Canada between 1952 and 2007. 

Here's an example that uses the `geom_point` from earlier...change this to a line.


**plot twist**

(See what I did there?)

You can add a `ggplot()` call as part of a pipe! This means that you can filter, mutate, group, etc and then pipe those results into a plot.

Note that the data is already identified, so you don't include it after `ggplot`.

And as we saw earlier, the `aes()` could also be included in the `geom_()` function.


```{r line3_1_1, exercise=TRUE, exercise.eval=FALSE}
# example: a point graph
gapminder %>%
  filter(country == "Canada") %>%
  # now the ggplot 
  ggplot(aes(x = year, y = lifeExp)) +  # IMPORTANT - it's a "+", not a "%>%"
  geom_point()


```



```{r line3_1_2, exercise=TRUE, exercise.eval=FALSE}

```

```{r line3_1_2-solution}


# solution 1
gapminder %>%
  filter(country == "Canada") %>%
  # now the ggplot 
  ggplot(aes(x = year, y = lifeExp)) +  # IMPORTANT - it's a "+", not a "%>%"
  geom_line()

# solution 2
gapminder %>%
  filter(country == "Canada") %>%
  # now the ggplot 
  ggplot() +  # IMPORTANT - it's a "+", not a "%>%"
  geom_line(aes(x = year, y = lifeExp))


```


### Your turn  3.2

Modify the code you wrote in #3.1 to show the life expectancy for Canada, Oman, and Japan. You will still have year and life expectancy as the point value represented on the axes -- how will you separate the countries?

(In 2007, Japan had the longest life expectancy of any country, and in the years between 1952 and 2007, Oman had the biggest increase--a change of 38 years! The code for how I determined this is at the bottom of the page.) 

```{r line3_2_1, exercise=TRUE, exercise.eval=FALSE}

```

```{r line3_2_1-solution}

country_list <- c("Canada", "Oman", "Japan")

# solution
gapminder %>%
  filter(country %in% country_list) %>%
  # now the ggplot 
  ggplot(aes(x = year, y = lifeExp, colour = country)) +  # IMPORTANT - it's a "+", not a "%>%"
  geom_line()


```

### Your turn  3.3

Layering geoms ... it is possible to add geoms to a plot. 

See what happens when you use the code from 3.2, with `geom_line`, and then add the `geom_point` from the example in 3.1.

```{r line3_2_1, exercise=TRUE, exercise.eval=FALSE}

```

```{r line3_2_1-solution}

# solution
gapminder %>%
  filter(country %in% c("Canada", "Oman", "Japan")) %>%
  # now the ggplot 
  ggplot(aes(x = year, y = lifeExp, colour = country)) +  # IMPORTANT - it's a "+", not a "%>%"
  geom_line() + 
  geom_point()


```













