---
title: "BIDA302: Introduction to R"
subtitle: "hands-on examples, with answers"
output: learnr::tutorial
runtime: shiny_prerendered
---



<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)
-->


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)

```

## Objectives

At the end of this tutorial, you will have:

* An understanding of and ability to use mathematical and assignment operators in R

* The ability to use basic statistical functions in R


## A. R as a calculator

In the code window below, type `10 + 4`, then press the "Run Code" button:

```{r A1, exercise=TRUE, exercise.eval=FALSE}



```

(The `[1]` is telling us that the "14", the result of our addition, is the first element in our vector.)



### other operators

In addition to these functions, R has a number of operators that add significantly to your code.

**Arithmetic operators**

| Operator | Description |
| :--- | :--- |
| + | *plus* |
| - | *minus* |
| \* | *multiplication*  |
| / | *division* |
| ^ | *exponential* |

<P>


### Your Turn A-1: math

Try some other math in the code window below:

```{r A1_yt, exercise=TRUE, exercise.eval=FALSE}



```



## B. The power of named variables 

In R, as in other programming languages, we can create named _objects_, and those objects can used in subsequent calculations. 


### the assignment arrow

In R, we use an assignment arrow to create a named object.

It looks like this:  `<-` 

It is a combination of a "less than" sign, followed by a hyphen. 


You can read it as:

`object_name <- value`

or even:

`new_thing <- old_thing`


In the example below, we will create an object "x" with the value 10. The object is then part of the R environment, and can be used later. 


To show `x` in the console, we type `x`

```{r B1, exercise=TRUE, exercise.eval=FALSE}

x <- 10

x

```




### Your Turn B.1: naming objects

Create a new object "y" with the value 4, and print it.

```{r B1-1_yt, exercise=TRUE, exercise.eval=FALSE}


```


```{r B1-1_yt-solution}
# answer
y <- 4

y

```


In the chunk below two objects are created and added together.


```{r B1-2, exercise=TRUE, exercise.eval=FALSE}

x <- 10

y <- 4

x + y

```



How would you set `x` equal to 20?

One solution:

`x <- 20`


Another solution: since we earlier assigned `x` a value of 10

`x <- x * 2`


Try the second solution in the chunk below:

```{r B1-3, exercise=TRUE, exercise.eval=FALSE}

x <- 10


```


```{r B1-3-solution}
# answer
x <- 10

x <- x * 2

x

```




### Your Turn B.2: more math


Create an object "x" with the value of 12, an object "y" with the value of 3, and an object "z" with a value of 7.  

Then divide "x" by "y", and "x" by "z".

Once you have your code written, run the chunk.

```{r B2-1_yt, exercise=TRUE, exercise.eval=FALSE}


```


```{r B2-1_yt-solution}
# answer
x <- 12
y <- 3
z <- 7

x/y

x/z

```



## C. Strings

In R, character strings are created by putting letters, numbers, and other symbols inside quotation marks:

`t <- "If it's inside quotation marks, it's a character"`


Let's print the object `hello_world`:

```{r C1, exercise=TRUE, exercise.eval=FALSE}

hello_world <- "Hello, world"


```

Note that if we put a number inside quotation marks, it's now a character string. If we try to use it like a number, R throws an error:

```{r C2_yt, exercise=TRUE, exercise.eval=FALSE}

not_a_number <- "10"

x <- 10

not_a_number + x 

```




## D. Dataframes

So far we've only looked at objects that have one item in them. Most of your data analysis is going to be in a data frame of some sort: think of a typical spreadsheet of numbers.

We will delve into good data frame structure soon, but for now let's practice with some small examples so you can see what's happening. In this example, we have a data frame with 2 variables, with 11 cases in each.



```{r D1}

anscombe_1 <- anscombe %>% 
  select(x1, y1)

```

```{r D1_yt, exercise=TRUE, exercise.setup = "D1"}

mean(anscombe_1$x1)


```

