---
title: "BIDA302: Introduction to R"
subtitle: "hands-on examples, with answers"
output: learnr::tutorial
runtime: shiny_prerendered
---



<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)
-->


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)

```

## Objectives

At the end of this tutorial, you will have:

* An understanding of and ability to use mathematical and assignment operators in R

* The ability to use basic statistical functions in R


## A. R as a calculator

In the code window below, type `10 + 4`, then press the "Run Code" button:

```{r A1, exercise=TRUE, exercise.eval=FALSE}



```

(The `[1]` is telling us that the "14", the result of our addition, is the first element in our vector.)



### other operators

In addition to these functions, R has a number of operators that add significantly to your code.

**Arithmetic operators**

| Operator | Description |
| :--- | :--- |
| + | *plus* |
| - | *minus* |
| \* | *multiplication*  |
| / | *division* |
| ^ | *exponential* |

<P>


### Your Turn A-1: math

Try some other math in the code window below:

```{r A1_yt, exercise=TRUE, exercise.eval=FALSE}



```



## B. The power of named variables 

In R, as in other programming languages, we can create named _objects_, and those objects can used in subsequent calculations. 


### the assignment arrow

In R, we use an assignment arrow to create a named object.

It looks like this:  `<-` 

It is a combination of a "less than" sign, followed by a hyphen. 


You can read it as:

`object_name <- value`

or even:

`new_thing <- old_thing`


In the example below, we will create an object "x" with the value 10. The object is then part of the R environment, and can be used later. 


To show `x` in the console, we type `x`

```{r B1, exercise=TRUE, exercise.eval=FALSE}

x <- 10

x

```




### Your Turn B.1: naming objects

Create a new object "y" with the value 4, and print it.

```{r B1-1_yt, exercise=TRUE, exercise.eval=FALSE}


```


```{r B1-1_yt-solution}
# answer
y <- 4

y

```


In the chunk below two objects are created and added together.


```{r B1-2, exercise=TRUE, exercise.eval=FALSE}

x <- 10

y <- 4

x + y

```



How would you set `x` equal to 20?

One solution:

`x <- 20`


Another solution: since we earlier assigned `x` a value of 10

`x <- x * 2`


Try the second solution in the chunk below:

```{r B1-3, exercise=TRUE, exercise.eval=FALSE}

x <- 10


```


```{r B1-3-solution}
# answer
x <- 10

x <- x * 2

x

```




### Your Turn B.2: more math


Create an object "x" with the value of 12, an object "y" with the value of 3, and an object "z" with a value of 7.  

Then divide "x" by "y", and "x" by "z".

Once you have your code written, run the chunk.

```{r B2-1_yt, exercise=TRUE, exercise.eval=FALSE}


```


```{r B2-1_yt-solution}
# answer
x <- 12
y <- 3
z <- 7

x/y

x/z

```



## C. Strings

In R, character strings are created by putting letters, numbers, and other symbols inside quotation marks:

`t <- "If it's inside quotation marks, it's a character"`


Let's print the object `hello_world`:

```{r C1, exercise=TRUE, exercise.eval=FALSE}

hello_world <- "Hello, world"


```

Note that if we put a number inside quotation marks, it's now a character string. If we try to use it like a number, R throws an error:

```{r C2_yt, exercise=TRUE, exercise.eval=FALSE}

not_a_number <- "10"

x <- 10

not_a_number + x 

```

## D. Lists & functions

The earlier examples have a single value in each object. Most data analysis is the summarization of multiple values—means, medians, correlations, and advanced machine learning methods are all approaches that take multiple values as inputs.

In R, one way to create one type of list (technically, an atomic vector) of numbers is using the `c()` function ("c" for "combine"). Here's the code to create a list of numbers:


```{r D1-setup}
# setup chunk with bida_numbers
bida_numbers <- c(301, 302, 404, 405, 406)

```

```{r D1-1, exercise=TRUE}

bida_numbers <- c(301, 302, 404, 405, 406)

bida_numbers

```

We can then apply a function to the object `bida_numbers`. For example, to calculate the mean, we use the function `mean()`:

```{r D1-2, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "D1-setup"}

mean(bida_numbers)

```

### Some summary functions


Some basic statistical functions  


**Two-variable statistical functions**

| Operator | Description |
| :--- | :--- |
| `min()` <br> `max()` | *minimum and maximum values* |
| `mean() <br> `median()` <br> `mode()` | *measures of "average"* |
| `sum()`  | *adds up the values* |
| `sd()`  | *calculates the standard deviation* |
| `var()`  | *calculates the variance* |
| `length()`  | *number of cases* |
| `summary()`  | *a list of measures of central tendency* |

<P>

Because R is a statistical/data science programming language, these and many more functions are built in to base R.


### Your Turn D.1: functions

Write the script to calculate:

* the minimum value of `bida_numbers`

* how many numbers there are in `bida_numbers`


```{r D1-3_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "D1-setup"}


```



```{r D1-3_yt-solution, exercise.setup = "D1-setup"}
# answer

min(bida_numbers)

length(bida_numbers)

```



### Your Turn D.2: `summary()` 

The `summary()` function is handy for getting a quick snapshot of a variable.


Write the script to generate the summary of `bida_numbers`.



```{r D2-1_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "D1-setup"}


```



```{r D2-1_yt-solution, exercise.setup = "D1-setup"}
# answer

summary(bida_numbers)

```


## E. Data frames

So far we've only looked at objects that have one item in them. But most of your data analysis is going to use a data frame (also known as a data table) of some sort think of a typical spreadsheet of numbers, with the data in columns and rows.

We will delve into what makes a good data frame structure later in this course, but for now let's practice with some small examples so you can see what's happening. 

In this example, we have a data frame called "anscombe_1" with 2 variables, and each variable has 11 cases in each. Let's take a quick look so we get a sense of the structure:



```{r E1}

anscombe_1 <- anscombe %>% 
  select(x1, y1)

```



```{r E1-1, exercise=TRUE, exercise.setup = "E1"}

anscombe_1

```


We can calculate the mean value of `x1` using the following code. The elements in the code are:

`function(dataframe$variable)`

Note that the dollar sign separates the dataframe from the variable in that dataframe. (Later in the course we will see other ways to write the same code.)


```{r E1-2, exercise=TRUE, exercise.setup = "E1"}

mean(anscombe_1$x1)


```


### Your Turn E.1: functions on data frames

Write the script to calculate:

* the minimum value of `x1`

* the median value of `y1`


```{r E1-3_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "E1"}


```



```{r E1-3_yt-solution, exercise.setup = "E1"}
# answer

min(anscombe_1$x1)

median(anscombe_1$y1)

```



### Some two-variable statistical functions

Earlier we saw functions that calculate summary statistics on a single variable, such as `mean()` and `sd()`. R also gives us functions to calculate the statistics on two variables, such as the correlation and the covariance.


**Two-variable statistical functions**

| Operator | Description |
| :--- | :--- |
| `cor(x, y)`  | *calculate the correlation between x and y* |
| `cov(x, y)`  | *calculate the covariance between x and y* |

<P>

And again, the statistical/data science foundation of R means that these and many more functions are built in to base R.


### Your Turn E.2: functions on two variables

Write the script to calculate:

* the correlation between `x1` and `y1`


```{r E2-1_yt, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "E1"}


```



```{r E2-1_yt-solution, exercise.setup = "E1"}
# answer

cor(anscombe_1$x1, anscombe_1$y1)

```

