---
title: "BIDA302: Introduction to R"
subtitle: "hands-on examples, with answers"
output: learnr::tutorial
runtime: shiny_prerendered
---



<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License

The tables of functions and operators are from Gina Reynolds, [Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)
-->


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

# packages used in the tutorial
library(tidyverse)
library(gapminder)

```

## Objectives

At the end of this tutorial, you will have:

* An understanding of and ability to use mathematical and assignment operators in R

* The ability to use basic statistical functions in R


## R as a calculator

In the code window below, type `10 + 4`, then press the "Run Code" button:

```{r calculator_example, exercise=TRUE, exercise.eval=FALSE}



```



### other operators

In addition to these functions, R has a number of operators that add significantly to your code.

**Arithmetic operators**

| Operator | Description |
| :--- | :--- |
| + | *plus* |
| - | *minus* |
| \* | *multiplication*  |
| / | *division* |
| ^ | *exponential* |
| \>    | *greater than, tests greater than* (also `>=`) |
| <     | *less than, tests less than* (also `<=`) |

<P>


## Your Turn 1.1: math

Try some other math in the code window below:

```{r calculator_freeform, exercise=TRUE, exercise.eval=FALSE}



```



## The power of named variables 

In R, as in other programming languages, various things get called _objects_ which can be named. Those names can then be written to use those named objects again. 

### the assignment arrow

In R, we use an assignment arrow to create a named object.

`<-`

> type "less than" plus hyphen


You can read it as:

`new_thing <- old_thing`


In the example below, we will create an object "x" with the value 10. The object is then part of the R environment, and can be used later. 


To show `x` in the console, we type `x`

```{r assignment-1-x, exercise=TRUE, exercise.eval=FALSE}

x <- 10

x

```


## Your Turn 1.2: naming objects

Create a new object "y" with the value 4

```{r assignment-2-y, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "assignment-1-x}


```


```{r assignment-2-y-solution}
# answer
y <- 4

y

```


> Now add x and y

```{r assignment_3_xy, exercise=TRUE, exercise.eval=FALSE, exercise.setup = "assignment_2_y"}

x + y

```



How would you set `x` equal to 20?

One solution:

`x <- 20`


Another solution: since we earlier assigned `x` a value of 10

`x <- x * 2`



And now divide `x` by `y`

`x / y`


`t <- "If it's inside quotation marks, it's a character"`


Now print the object `t`

